---
title: "validationAnalysis - "
author: "Sophie Seidel"
date: "1/25/2021"
output: pdf_document
---

This is well-calibrated simulation study, where the following parameters are sampled from the prior:

 - birth rate
 - sampling proportion
 - scarring rates
 
The death rate, the origin, the time interval of scarring are fixed.

For all parameters, the prior used to sample parameters from for the simulation and the prior used in the inference are identical.


```{r setup}
rm(list = ls())
knitr::opts_chunk$set(echo = FALSE)
source("scripts.R")
source("inference_performance.R")
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
```


```{r }
library(tracerer)
library(HDInterval)
library(ggplot2)
```

```{r}
thisAnalysis = "realistic_inference_1type"
N = 1000
resultsDir = "../inference/inferenceOutput/"
simulationDir = "../simulation/simParams/"
simulationOutputDir = "../simulation/simulationOutput"
#simFiles = list.files(path = simulationDir, pattern = "*.csv", )
simOutFiles = list.files(path = simulationOutputDir, pattern = "*log")
#plotDir = "~/Projects/trees-in-devBio/results/packageDev/inSilicoValidation/fixPhylodynamicParams/plots/"
metricDir = "../analysis/metrics/"
```




```{r}
parameters = c("treeHeight", "treeLength", "birthRate", "deathRate", "rho", paste0("scarringRate.", 1:50), "clockRate")
nParams = length(parameters)
exampleDat = read.csv(file = paste0(resultsDir, "realistic_inference_1type.1.log"), header = TRUE, sep = "\t")

parameter_col_index_map = list(6, 7, 59, 61, 60, 8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23,24,25,26,27,28,29,30,31,32,33,34,35,36,37,38,39,40,41,42,43,44,45,46,47,48,49,50,51,52,53,54,55,56,57, 
                               62)
names(parameter_col_index_map) = parameters
```

```{r ouputData}
outputDat = data.frame(matrix(nrow = N,  ncol = 3*length(parameters) + 1))
colnames(outputDat) = c(get_sumstat_names_from_parameters(parameters), "seed")


simParams = as.data.frame(matrix(nrow = 1000, ncol = nParams+3, data = 0))
colnames(simParams) = c(parameters, "scarPresent", "nCells", "seed")

```




```{r }
#load(paste(resultsDir, thisAnalysis, ".Rdat", sep = ""))
files = list.files(resultsDir, pattern = "*.log")
 
for (ctr in 1:1000){
     # get log file and its simulation seed
    print(ctr)
    #exclude tree with >15000 leafs
    if(ctr == 874){
    next()
    }
    #file = files[ctr]
    #seed = get_seed_from_file_name(file)

    seed = ctr  
    # parse beast log and save summary stats in output data frame
    # log =  parse_beast_tracelog_file(paste(resultsDir, file, sep = "/"))
    # log[, c(8:57)] = log[, c(8:57)] * log[, 62] # multiply scarring rates with clock rate
    # outputDat[ctr, 1:(nParams*3)] = get_summary_stats_from_log(log,parameter_col_index_map =parameter_col_index_map)
    # outputDat[ctr, "seed"] = seed
    
    # parse simulation files and save true simulation params in simParams data frame
  simFile =  paste0(simulationDir, "simParams_", seed, ".csv" ) #simFiles[f]
  simParams[ctr, c(paste0("scarringRate.", 1:50),  "clockRate", "birthRate", "deathRate", "rho")] =
    get_simulation_params_from_file(simFile) 
  alignment = paste0(simulationOutputDir, "/realistic_simulation_1type.", seed, ".alignment.nexus")
  simParams[ctr, "scarPresent"] = scar_present(alignment)
  simParams[ctr, "nCells"] = as.numeric(get_ntaxa(alignment))

  simLog = paste0(simulationOutputDir, 
  "/realistic_simulation_1type.", seed,   ".log")
  simParams[ctr, c("treeHeight", "treeLength")] = get_tree_params_from_sim_log(simLog)
  simParams[ctr, "seed"] = seed
  
}
outputDat = outputDat[order(as.numeric(outputDat$seed)), ]
outputDat$seed = as.numeric(outputDat$seed)
simParams = simParams[order(as.numeric(simParams$seed)), ]

save(outputDat, file = paste(resultsDir, thisAnalysis, ".Rdat", sep = ""))
simParams[, 6:55] = simParams[, 6:55] * simParams[, 56] # multiply scarring rates with clock rate
save(simParams, file = paste(resultsDir, thisAnalysis, "_simParams.Rdat", sep = ""))
```


```{r}
load(file = paste(resultsDir, thisAnalysis, ".Rdat", sep = ""))
load(file = paste(resultsDir, thisAnalysis, "_simParams.Rdat", sep = ""))
```

```{r}
exclude_seeds = c()
outputDat_cleaned = outputDat[which(!(outputDat$seed %in% exclude_seeds)), ]
simParams_cleaned = simParams[which(!(simParams$seed %in% exclude_seeds)), ]
simParams_cleaned = simParams_cleaned[which(simParams_cleaned$seed %in% outputDat_cleaned$seed), ]

#remove chains that have not run
outputDat_cleaned = outputDat_cleaned[which(outputDat_cleaned$treeHeight_lower != -1), ]
outputDat_cleaned = unique(outputDat_cleaned)
simParams_cleaned = simParams_cleaned[which(simParams_cleaned$treeHeight != -1), ]
```


#check which chains have converged
```{r}
seeds = outputDat_cleaned$seed
max(simParams[which(simParams$seed %in% seeds), "nCells"])
notConverged = simParams[which(!(simParams$seed %in% seeds)), ]
notConverged = notConverged[2:nrow(notConverged), ] # remove first placeholder row
notConverged = notConverged[order(notConverged$nCells), ]
notConverged[1:14, c("nCells", "seed")]
```


```{r}
seeds_wo_tree = simParams_cleaned[which(simParams_cleaned$treeHeight == 0), "seed"]
# outputDat_wo_tree = outputDat_cleaned[which(outputDat_cleaned$seed %in% seeds_w_tree), ]
# outputDat_w_tree = outputDat_cleaned[which(!(outputDat_cleaned$seed %in% seeds_w_tree)), ]
outputDat_cleaned$wTree = "> 1 cell"
outputDat_cleaned[which(outputDat_cleaned$seed %in% seeds_wo_tree), "wTree"] = "1 cell"
simParams_cleaned$wTree = outputDat_cleaned$wTree
```

```{r}
outputDat_cleaned$scarPresent = simParams_cleaned$scarPresent
outputDat_cleaned$scar = "scar present"
outputDat_cleaned[which(!(outputDat_cleaned$scarPresent)), "scar"] = "scar not present"
simParams_cleaned$scar = outputDat_cleaned$scar
#outputDat_cleaned[which(!(simParams_cleaned$scarPresent)), "scarPresent"] = "no scar present"
#simParams_cleaned$scarPresent = outputDat_cleaned$scarPresent
```

# describe simulated data

```{r}
ggplot(simParams, aes(x=nCells))+ geom_histogram() + xlim(0,1000)
```



# Bias



```{r}
bias_all = get_bias_from_dat(outputDat = outputDat_cleaned, simParams = simParams_cleaned, parameters = parameters)
n = nrow(outputDat_cleaned)
meanbias = apply(X = bias_all, MARGIN = 2, function(x){mean(x)})
sdbias = apply(X = bias_all, MARGIN = 2, function(x){sd(x)/sqrt(n)})

saveRDS(rbind(meanbias, sdbias), file = paste0(metricDir, "bias_all.Rds"))
```


```{r}
outputDat_first100 = outputDat_cleaned[which(outputDat_cleaned$seed <= 100), ]
simParams_first100 = simParams_cleaned[which(simParams_cleaned$seed <=100), ]
n_f100 = nrow(outputDat_first100)
bias_first100 = get_bias_from_dat(outputDat = outputDat_first100, simParams = simParams_first100, parameters = parameters)
meanbias_f100 = apply(X = bias_first100, MARGIN = 2, function(x){mean(x)})
sdbias_f100 = apply(X = bias_first100, MARGIN = 2, function(x){sd(x)/sqrt(n_f100)})

saveRDS(rbind(meanbias_f100, sdbias_f100), file = paste0(metricDir, "bias.Rds"))
```


#RMSE



```{r}
rmse_all = get_rmse_from_dat(outputDat = outputDat_cleaned, simParams = simParams_cleaned, parameters = parameters)

meanrmse = apply(X = rmse_all, MARGIN = 2, function(x){mean(x)})
sdrmse = apply(X = rmse_all, MARGIN = 2, function(x){sd(x)/sqrt(n)})

saveRDS(rbind(meanrmse, sdrmse), file = paste0(metricDir, "rmse_all.Rds"))
```

```{r}
rmse_first100 = get_rmse_from_dat(outputDat = outputDat_first100, simParams = simParams_first100, parameters = parameters)
meanrmse_f100 = apply(X = rmse_first100, MARGIN = 2, function(x){mean(x)})
sdrmse_f100 = apply(X = rmse_first100, MARGIN = 2, function(x){sd(x)/sqrt(n_f100)})
saveRDS(rbind(meanrmse_f100, sdrmse_f100), file = paste0(metricDir, "rmse.Rds"))
```

#HPD



```{r}
hpd_all = get_hpd_from_dat(outputDat_cleaned, simParams_cleaned, parameters)
meanhpd = apply(X = hpd_all, MARGIN = 2, function(x){mean(x)})
sdhpd = apply(X = hpd_all, MARGIN = 2, function(x){sd(x)/sqrt(n)})
saveRDS(rbind(meanhpd, sdhpd), file = paste0(metricDir, "hpd_all.Rds"))
```

```{r}
hpd_first100 = get_hpd_from_dat(outputDat_first100, simParams_first100, parameters)
meanhpd_f100 = apply(X = hpd_first100, MARGIN = 2, function(x){mean(x)})
sdhpd_f100 = apply(X = hpd_first100, MARGIN = 2, function(x){sd(x)/sqrt(n_f100)})
saveRDS(rbind(meanhpd_f100, sdhpd_f100), file = paste0(metricDir, "hpd.Rds"))
```

# Coverage



```{r}
coverage_all = get_coverage_from_dat(outputDat_cleaned, simParams_cleaned, parameters)
meancov = apply(X = coverage_all, MARGIN = 2, function(x){mean(x)})
sdcov = apply(X = coverage_all, MARGIN = 2, function(x){sd(x)/sqrt(n)})
saveRDS(rbind(meancov, sdcov), file = paste0(metricDir, "coverage_all.Rds"))
```

```{r}
coverage_first100 = get_coverage_from_dat(outputDat_first100, simParams_first100, parameters)
meancov_f100 = apply(X = coverage_first100, MARGIN = 2, function(x){mean(x)})
sdcov_f100 = apply(X = coverage_first100, MARGIN = 2, function(x){sd(x)/sqrt(n_f100)})
saveRDS(rbind(meancov_f100, sdcov_f100), file = paste0(metricDir, "coverage.Rds"))
```


```{r}
sim_s700 = simParams[which(simParams$nCells < 700), ]
summary(sim_s700)
```


To enter into supp table on validation metrics. Don't report scarring rate 50 because it was fixed to 1 and don't report clock rate as we report the product of each scarring rate with the clock rate.
```{r}
metrics = data.frame(parameter=names(meancov), coverage = meancov, hpd= meanhpd, rmse=meanrmse, bias=meanbias)
metrics[ ,2:5] = round(metrics[,2:5], digits = 2)
rows = apply(metrics, 1, function(x){paste(x, collapse = " & ")})
paste(rows, collapse = " \\")
```
# Quantify correlations

```{r}
combined = merge(outputDat_cleaned, simParams_cleaned, by = "seed")
```


Treeheight

```{r}
cor.test(combined$treeHeight, combined$treeHeight_median)
```

```{r}
poster_width_ab16 = combined[which(combined$treeHeight >= 16), c("treeHeight_upper", "treeHeight_lower", 
                                                                 "treeHeight_median", "treeHeight")]
poster_width_bel16 = combined[which(combined$treeHeight < 16), c("treeHeight_upper", "treeHeight_lower", 
                                                                 "treeHeight_median","treeHeight")]

cor.test(poster_width_ab16$treeHeight_median, poster_width_ab16$treeHeight)
cor.test(poster_width_bel16$treeHeight_median, poster_width_bel16$treeHeight)


poster_width_ab16= mean( (poster_width_ab16$treeHeight_upper - poster_width_ab16$treeHeight_lower)/poster_width_ab16$treeHeight)
poster_width_bel16= mean( (poster_width_bel16$treeHeight_upper - poster_width_bel16$treeHeight_lower)/poster_width_bel16$treeHeight, na.rm = T)

poster_width_ab16
poster_width_bel16
```

tree length

```{r}
cor.test(combined$treeLength_median, combined$treeLength)
```

birth rate
```{r}
cor.test(combined$birthRate_median, combined$birthRate)
```

rho
```{r}
cor.test(combined$rho, combined$rho_median)
```


<!-- # Posterior intervals  -->

<!-- ## for tree height -->

<!-- ```{r} -->
<!-- ordering = order(simParams_cleaned$treeHeight) -->
<!-- simParams_cleaned = simParams_cleaned[ordering, ] -->
<!-- simParams_cleaned$orderedSeed = 1:nrow(outputDat_cleaned) -->
<!-- outputDat_cleaned = outputDat_cleaned[ordering, ] -->
<!-- outputDat_cleaned$orderedSeed = 1:nrow(outputDat_cleaned) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- p = ggplot(data = outputDat_cleaned, aes(x=as.numeric(orderedSeed), y=treeHeight_median)) +  -->
<!--   #facet_grid(wTree~scar)+ -->
<!--   geom_point() + -->
<!--   geom_errorbar(aes(ymin=treeHeight_lower, ymax=treeHeight_upper), alpha=0.4) + -->
<!--   geom_point(data = simParams_cleaned, aes(x=as.numeric(orderedSeed), y=treeHeight, size="truth"), inherit.aes = FALSE, col="darkgreen", size=0.3) + -->
<!--   theme_minimal()+ -->
<!--   theme()+ -->
<!--   geom_hline(yintercept = 16, linetype = "dashed", show.legend = TRUE)+ -->
<!--   xlab("Ordered simulation number")+ -->
<!--   ylab("Tree height") -->
<!-- ggsave(plot = p, filename = paste0(plotDir, "/posteriorIntervals_treeHeight_", thisAnalysis, ".pdf"), width = 8, height = 6, units = "cm") -->
<!-- p -->

<!-- ``` -->

<!-- ## for tree length -->

<!-- ```{r} -->
<!-- ordering = order(simParams_cleaned$treeLength) -->
<!-- simParams_cleaned = simParams_cleaned[ordering, ] -->
<!-- simParams_cleaned$orderedSeed = 1:nrow(outputDat_cleaned) -->
<!-- outputDat_cleaned = outputDat_cleaned[ordering, ] -->
<!-- outputDat_cleaned$orderedSeed = 1:nrow(outputDat_cleaned) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- p <- ggplot(data = outputDat_cleaned, aes(x=as.numeric(orderedSeed), y=treeLength_median)) +  -->
<!--   geom_point() + -->
<!--   geom_errorbar(aes(ymin=treeLength_lower, ymax=treeLength_upper), alpha=0.4) + -->
<!--   geom_point(data = simParams_cleaned, aes(x=as.numeric(orderedSeed), y=treeLength, size="truth"), inherit.aes = FALSE, col="darkgreen", size=0.3) + -->
<!--   theme_minimal()+ -->
<!--   theme(legend.title = element_blank())+ -->
<!--   xlab("Ordered simulation number")+ -->
<!--   ylab("Posterior intervals for tree length") -->

<!-- ggsave(plot = p, filename = paste0(plotDir, "/posteriorIntervals_treeLength_", thisAnalysis, ".pdf")) -->

<!-- p -->

<!-- ``` -->

<!-- ## for scarring rate 1 -->

<!-- ```{r} -->
<!-- ordering = order(simParams_cleaned$scarringRate.1) -->
<!-- simParams_cleaned = simParams_cleaned[ordering, ] -->
<!-- simParams_cleaned$orderedSeed = 1:nrow(outputDat_cleaned) -->
<!-- outputDat_cleaned = outputDat_cleaned[ordering, ] -->
<!-- outputDat_cleaned$orderedSeed = 1:nrow(outputDat_cleaned) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- p <- ggplot(data = outputDat_cleaned, aes(x=as.numeric(orderedSeed), y=scarringRate.1_median)) +  -->
<!--   geom_point() + -->
<!--   geom_errorbar(aes(ymin=scarringRate.1_lower, ymax=scarringRate.1_upper), alpha=0.4) + -->
<!--   geom_point(data = simParams_cleaned, aes(x=as.numeric(orderedSeed), y=scarringRate.1, size="truth"), inherit.aes = FALSE, col="darkgreen", size=0.3) + -->
<!--   theme_minimal()+ -->
<!--   theme(legend.title = element_blank())+ -->
<!--   xlab("Ordered simulation number")+ -->
<!--   ylab("Posterior intervals for scarring rate 1") -->

<!-- ggsave(plot = p, filename = paste0(plotDir, "/posteriorIntervals_scarringRate1_", thisAnalysis, ".pdf")) -->

<!-- p -->

<!-- ``` -->



<!-- ## for clock rate -->

<!-- ```{r} -->
<!-- ordering = order(simParams_cleaned$clockRate) -->
<!-- simParams_cleaned = simParams_cleaned[ordering, ] -->
<!-- simParams_cleaned$orderedSeed = 1:nrow(outputDat_cleaned) -->
<!-- outputDat_cleaned = outputDat_cleaned[ordering, ] -->
<!-- outputDat_cleaned$orderedSeed = 1:nrow(outputDat_cleaned) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- p <- ggplot(data = outputDat_cleaned, aes(x=as.numeric(orderedSeed), y=clockRate_median)) +  -->
<!--   geom_point() + -->
<!--   geom_errorbar(aes(ymin=clockRate_lower, ymax=clockRate_upper), alpha=0.4) + -->
<!--   geom_point(data = simParams_cleaned, aes(x=as.numeric(orderedSeed), y=clockRate, size="truth"), inherit.aes = FALSE, col="darkgreen", size=0.3) + -->
<!--   theme_minimal()+ -->
<!--   theme(legend.title = element_blank())+ -->
<!--   xlab("Ordered simulation number")+ -->
<!--   ylab("Posterior intervals for clock rate") -->

<!-- ggsave(plot = p, filename = paste0(plotDir, "/posteriorIntervals_clockRate_", thisAnalysis, ".pdf")) -->

<!-- p -->

<!-- ``` -->

<!-- ## for death rate  -->
<!-- ```{r} -->
<!-- ordering = order(outputDat_cleaned$deathRate_median) -->
<!-- simParams_cleaned = simParams_cleaned[ordering, ] -->
<!-- simParams_cleaned$orderedSeed = 1:nrow(outputDat_cleaned) -->
<!-- outputDat_cleaned = outputDat_cleaned[ordering, ] -->
<!-- outputDat_cleaned$orderedSeed = 1:nrow(outputDat_cleaned) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- p = ggplot(data = outputDat_cleaned, aes(x=as.numeric(orderedSeed), y=deathRate_median)) +  -->
<!--   geom_point() + -->
<!--   geom_errorbar(aes(ymin=deathRate_lower, ymax=deathRate_upper), alpha=0.4) + -->
<!--   geom_hline(yintercept = 0.01, linetype = "dashed") + geom_hline(yintercept = 1.4, linetype = "dashed")+  -->
<!--   geom_hline(yintercept = 0.4, col = "darkgreen")+ -->
<!--   theme_minimal()+ -->
<!--   theme(legend.title = element_blank())+ -->
<!--   xlab("Ordered simulation number")+ -->
<!--   ylab("Posterior intervals for death rate") -->
<!-- ggsave(plot = p, filename = paste0(plotDir, "/posteriorIntervals_deathRate_", thisAnalysis, ".pdf")) -->
<!-- p -->
<!-- ``` -->
<!-- ## for rho -->


<!-- ```{r} -->
<!-- ordering = order(simParams_cleaned$rho) -->
<!-- simParams_cleaned = simParams_cleaned[ordering, ] -->
<!-- simParams_cleaned$orderedSeed = 1:nrow(outputDat_cleaned) -->
<!-- outputDat_cleaned = outputDat_cleaned[ordering, ] -->
<!-- outputDat_cleaned$orderedSeed = 1:nrow(outputDat_cleaned) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- p = ggplot(data = outputDat_cleaned, aes(x=as.numeric(orderedSeed), y=rho_median)) +  -->
<!--   facet_grid(wTree~scar)+ -->
<!--   geom_point() + -->
<!--   geom_errorbar(aes(ymin=rho_lower, ymax=rho_upper), alpha=0.4) + -->
<!--   geom_point(data = simParams_cleaned, aes(x=as.numeric(orderedSeed), y=rho, size="truth"), inherit.aes = FALSE, col="darkgreen") + -->
<!--   theme_minimal()+ -->
<!--   theme(legend.title = element_blank())+ -->
<!--   geom_hline(yintercept = 0.006)+ -->
<!--   geom_hline(yintercept = 0.31)+ -->
<!--   xlab("Ordered simulation number")+ -->
<!--   ylab("Posterior intervals for rho") -->
<!-- ggsave(plot = p, filename = paste0(plotDir, "/posteriorIntervals_rho_", thisAnalysis, ".pdf")) -->
<!-- p -->
<!-- ``` -->

<!-- ## for birth rate -->

<!-- ```{r} -->
<!-- ordering = order(simParams_cleaned$birthRate) -->
<!-- simParams_cleaned = simParams_cleaned[ordering, ] -->
<!-- simParams_cleaned$orderedSeed = 1:nrow(outputDat_cleaned) -->
<!-- outputDat_cleaned = outputDat_cleaned[ordering, ] -->
<!-- outputDat_cleaned$orderedSeed = 1:nrow(outputDat_cleaned) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- birthRatePrior = rlnorm(n = 10000, meanlog = -0.6, sdlog = 0.1) -->
<!-- quantiles_br = quantile(birthRatePrior, c(0.05, 0.95)) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- p = ggplot(data = outputDat_cleaned, aes(x=as.numeric(orderedSeed), y=birthRate_median)) +  -->
<!--   facet_grid(wTree~scar)+ -->
<!--   geom_point() + -->
<!--   geom_errorbar(aes(ymin=birthRate_lower, ymax=birthRate_upper), alpha=0.4) + -->
<!--   geom_point(data = simParams_cleaned, aes(x=as.numeric(orderedSeed), y=birthRate, size="truth"), inherit.aes = FALSE, col="darkgreen", size =0.3) + -->
<!--   theme_minimal()+ -->
<!--   geom_hline(yintercept = quantiles_br[1])+ -->
<!--   geom_hline(yintercept = quantiles_br[2]) + -->
<!--   theme(legend.title = element_blank())+ -->
<!--   xlab("Ordered simulation number")+ -->
<!--   ylab("Posterior intervals for birth rate") -->
<!-- ggsave(plot = p, filename = paste0(plotDir, "/posteriorIntervals_birthRate_", thisAnalysis, ".pdf")) -->
<!-- p -->
<!-- ``` -->


